Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of neutron star - black hole binary observations

### 2016-10-18 Paper/paper.tex
incorporated all of harald's comments but for fig10

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Techniques}\label{s1:techniques}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% #################
% THESE FIGURES ARE FOR THE NEXT SECTION
\begin{figure*}
\centering 
% \textbf{Illustrative posterior probability distributions for mass-ratio $\eta$,}\par
% \textbf{at different SNR values (shown in legend).}\par\medskip
\includegraphics[width=1.8\columnwidth]{plots/SingleSystemEta_q4_0_mc2_25_chi0_50}\\
\caption{{\bf Illustrative posterior probability distributions for mass-ratio $\eta$ at different SNR values:}
We show here probability distributions for mass ratio $\eta$ as measured
for the same signal at different SNRs. The intrinsic parameters of the source
are: $q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$, and $\lambdans=2000$;
and the signal is injected at SNRs $\rho=\{20,30,50\}$ (left to right). The templates
{\it ignore} tidal effects.
% 
In each panel: the dashed red line marks the median value
$\eta^\mathrm{Median}$, while the dashed green line show the true value
$\eta^\mathrm{Injected}$. The darker shading shows
the recovered $90\%$ credible interval for $\eta$, $(\Delta\eta)^{90\%}$.
% 
Comparing systematic and statistical errors, we find that:
at $\rho=20$, $\eta$ measurement is dominated by statistical
errors; at $\rho=30$, the two become comparable; and 
for louder signals ($\rho\simeq50$), the systematic errors dominate.
}
\label{fig:SingleSystemEtaPDFvsSNR}
\end{figure*}
% 
% ###############################
% In this section we summarize various technical aspects of this paper. The model
% used to obtain tidal waveforms is described in Sec.~\ref{s2:waveforms}. The 
% process of inferring source parameters from a GW signal is summarized in 
% Sec.~\ref{s2:bayesian}. The process of generating and combining multiple events
% is instead detailed in the corresponding results section,
% Sec.~\ref{s2:astro_multiple}.

@@

% \begin{figure*}
% \centering 
% \textbf{Illustrative posterior probability distributions for mass-ratio $\eta$,}\par
% \textbf{at different SNR values (shown in legend).}\par\medskip
% \includegraphics[width=1.8\columnwidth]{plots/SingleSystemEta_q4_0_mc2_25_chi0_50}
% \caption{We show here probability distributions for mass ratio $\eta$ as measured
% for the same injection at different SNRs. The intrinsic parameters of the source
% are: $q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$, and $\lambdans=2000$;
% and the signal is injected at SNRs $\rho=\{20,30,50\}$ (left to right). The templates
% {\it ignore} tidal effects.
% In each panel: the dashed red line marks the median value
% $\eta^\mathrm{Median}$, while the dashed green line show the true value
% $\eta^\mathrm{Injected}$. The difference between the true and median values
% is the systematic bias introduced
% by ignoring tidal effects in templates. The darker shading shows
% the recovered $90\%$ credible interval for $\eta$, $(\Delta\eta)^{90\%}$, 
% which is a direct measure of our statistical uncertainty on $\eta$.
% Comparing systematic and statistical errors, we find that:
% at $\rho=20$, $\eta$ measurement is dominated by statistical
% errors; at $\rho=30$, the two become comparable; and 
% for louder signals ($\rho\simeq50$), the systematic errors dominate.
% \label{fig:SingleSystemEtaPDFvsSNR}
% \end{figure*}
% ################# FIXME FIXME
% \begin{figure*}
% \centering 
% \textbf{Statistical measurement uncertainty for NSBH parameters $\mchirp$ (top),
% $\eta$ (middle) and $\chibh$ (bottom);\newline ignoring tidal effects.}\par\medskip
% \includegraphics[trim = {2cm 0 0 0},width=2.1\columnwidth]{plots/TNMchirpCIWidths90_0_Lambda_SNR}\\
% \includegraphics[trim = {2cm 0 0 0},width=2.1\columnwidth]{plots/TNEtaCIWidths90_0_Lambda_SNR}\\
% \includegraphics[trim = {2cm 0 0 0},width=2.1\columnwidth]{plots/TNChiBHCIWidths90_0_Lambda_SNR}
% \caption{
% We show here the statistical uncertainty of our measurement of the 
% dimensionless mass ratio $\eta$ (at $90\%$ credibility), over the NSBH
% parameter space.
% Individual panels show the same as a function of BH mass and spin.
% Across each row, we see the effect of increasing signal strength (i.e.
% SNR) with the tidal deformability of the NS $\lambdans$ fixed. Down each
% column, we see the effect of increasing $\lambdans$, at fixed SNR.
% \label{fig:CIWidths90_Lambda_SNR}
% \end{figure*}
% #################

@@

% \textbf{Ratio of systematic to statistical errors in measuring $\mchirp$;}\par
% \textbf{ignoring tidal effects.}\par\medskip
\includegraphics[width=1.95\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear}
% \includegraphics[width=0.8\columnwidth]{plots/TNMchirpBiasesOverCIWidthsVsSNR_Lambda500_CI90_0}
% % \includegraphics[width=0.67\columnwidth]{plots/TNMchirpBiasesOverCIWidthsVsSNR_Lambda800_CI90_0}
% \includegraphics[width=0.8\columnwidth]{plots/TNMchirpBiasesOverCIWidthsVsSNR_Lambda1000_CI90_0}\\
% \includegraphics[width=0.8\columnwidth]{plots/TNMchirpBiasesOverCIWidthsVsSNR_Lambda1500_CI90_0}
% \includegraphics[width=0.8\columnwidth]{plots/TNMchirpBiasesOverCIWidthsVsSNR_Lambda2000_CI90_0}

@@

% \textbf{Ratio of systematic to statistical errors in measuring $\eta$;}\par
% \textbf{ignoring tidal effects.}\par\medskip
\includegraphics[width=1.95\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR20_50_linear}
% \includegraphics[width=0.8\columnwidth]{plots/TNEtaBiasesOverCIWidthsVsSNR_Lambda500_CI90_0}
% % \includegraphics[width=0.67\columnwidth]{plots/TNEtaBiasesOverCIWidthsVsSNR_Lambda800_CI90_0}
% \includegraphics[width=0.8\columnwidth]{plots/TNEtaBiasesOverCIWidthsVsSNR_Lambda1000_CI90_0}\\
% \includegraphics[width=0.8\columnwidth]{plots/TNEtaBiasesOverCIWidthsVsSNR_Lambda1500_CI90_0}
% \includegraphics[width=0.8\columnwidth]{plots/TNEtaBiasesOverCIWidthsVsSNR_Lambda2000_CI90_0}

@@

% \textbf{Ratio of systematic to statistical errors in measuring $\chibh$;}\par
% \textbf{ignoring tidal effects.}\par\medskip
\includegraphics[width=1.95\columnwidth]{plots/TNChiBHBiasesOverCIWidths_CI90_0_Lambda_SNR_linear}
% % \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI68_3.pdf}
% % \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI90_0.pdf}
% \includegraphics[width=0.8\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_Lambda500_CI90_0}
% % \includegraphics[width=0.67\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_Lambda800_CI90_0}
% \includegraphics[width=0.8\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_Lambda1000_CI90_0}\\
% \includegraphics[width=0.8\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_Lambda1500_CI90_0}
% \includegraphics[width=0.8\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_Lambda2000_CI90_0}

@@

% \textbf{Illustrative posterior probability distributions for NS tidal
% deformability $\lambdans$,}\par
% \textbf{for signals with different $\lambdans$ values; $\lambdans = 1000,1500,2000,$
% from left to right.}\par\medskip
% \includegraphics[width=1.8\columnwidth]{plots/SingleSystem_q4_0_mc2_25_chi0_50}
\includegraphics[width=1.9\columnwidth]{plots/SingleSystemLambdaVary_q4_0_mc2_25_chi0_50_snr50}

@@

% Comparing systematic and statistical errors, we find that:
% at $\rho=20$, $\lambdans$ measurement is dominated by statistical
% errors; at $\rho=30$, the two become comparable; and 
% for louder signals ($\rho\simeq50$), the systematic errors dominate.

@@

\section{What do we gain by using templates that include NS matter effects?}\label{s1:PEwithNS}
% \begin{figure*}
% \centering 
% \textbf{Illustrative posterior probability distributions for NS tidal
% deformability $\lambdans$,}\par
% \textbf{for signals with different $\lambdans$ values; $\lambdans = 1000,1500,2000,$
% from left to right.}\par\medskip
% % \includegraphics[width=1.8\columnwidth]{plots/SingleSystem_q4_0_mc2_25_chi0_50}
% \includegraphics[width=1.9\columnwidth]{plots/SingleSystemLambdaVary_q4_0_mc2_25_chi0_50_snr50}
% \caption{We illustrate here probability distributions recovered for the NS tidal
% deformability parameter $\lambdans$ from three GW injections, with parameters:
% $q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$, and 
% $\lambdans=\{1000,1500,2000\}$ from left to right. The injection SNR is fixed at
% $\rho=50$. The templates {\it include} tidal effects, with a prior $0\leq\lambdans\leq 4000$.
% % These figures show that this approximation holds up to (SNR)
% % $\rho\simeq 30-50$.
% In each panel- the dashed red line marks the median value for
% $\lambdans$, and the dashed green line marks its {\it true} value.
% The darker shading shows the $90\%$ credible interval, whose width
% $(\Delta\lambdans)^{90\%}$ is a direct measure of our statistical uncertainty.
% By comparing the measurement uncertainty for these three injections, we see
% that $(\Delta\lambdans)^{90\%}$ grows very slowly with $\lambdans$. Therefore,
% the fractional measurement error - $(\Delta\lambdans)^{90\%}/\lambdans$ -
% decreases monotonically as $\lambdans$ increases (with signal strength fixed).
% % Comparing systematic and statistical errors, we find that:
% % at $\rho=20$, $\lambdans$ measurement is dominated by statistical
% % errors; at $\rho=30$, the two become comparable; and 
% % for louder signals ($\rho\simeq50$), the systematic errors dominate.
% \label{fig:SingleSystemLambdaPDFvsSNR}
% \end{figure*}
% % %% FIXME FIXME
% % \begin{figure*}
% % \centering
% % \textbf{Illustrative posterior probability distributions for NSBH parameters,}\par
% % \textbf{for signals at different SNR values; $\rho = 20$ (left) and $50$ (right).}\par\medskip
% % \includegraphics[width=1.05\columnwidth,trim=2cm 0 0 0]{plots/AllParamsMcEtPDF1D2D_q4_mc2_25_chi0_50_snr20}%\\
% % \includegraphics[width=1.05\columnwidth,trim=2cm 0 0 0]{plots/AllParamsMcEtPDF1D2D_q4_mc2_25_chi0_50_snr50}%\\
% % % \includegraphics[width=0.5\columnwidth]{plots/McEta2D_q4_mc2_25_chi0_50_snr30}
% % % \includegraphics[width=0.5\columnwidth]{plots/EtaChiBH2D_q4_mc2_25_chi0_50_snr30}
% % % \includegraphics[width=0.5\columnwidth]{plots/McChiBH2D_q4_mc2_25_chi0_50_snr30}\\
% % % \includegraphics[width=0.5\columnwidth]{plots/LambdaMc2D_q4_mc2_25_chi0_50_snr30}
% % % \includegraphics[width=0.5\columnwidth]{plots/LambdaEta2D_q4_mc2_25_chi0_50_snr30}
% % % \includegraphics[width=0.5\columnwidth]{plots/LambdaChiBH2D_q4_mc2_25_chi0_50_snr30}\\
% % % \includegraphics[width=0.47\columnwidth]{plots/LambdaMc2D_q4_mc2_25_chi0_50_snr50}
% % % \includegraphics[width=0.47\columnwidth]{plots/LambdaEta2D_q4_mc2_25_chi0_50_snr50}
% % % \includegraphics[width=0.47\columnwidth]{plots/LambdaChiBH2D_q4_mc2_25_chi0_50_snr50}
% % \caption{%
% % We illustrate here two-dimensional joint probability distributions, as
% % recovered for the intrinsic parameters of a binary with
% % $q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$,
% % $\lambdans=2000$, and $\rho=30$ (same as in
% % Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}).
% % The top row shows the same for combinations of 
% % non-tidal parameters, i.e. binary masses and spins, while the middle row 
% % shows combinations of $\lambdans$ with non-tidal parameters. The bottom row
% % is similar to the middle, with the only difference being that the SNR has been
% % turned up to $\rho=50$.
% % }
% % \label{fig:SingleSystemLambda2DPDFs}
% % \end{figure*}
% #################
% \textbf{Statistical uncertainty in $\lambdans$ measurement}\par\medskip
\includegraphics[trim={3cm 0 0 0.5cm},width=2.2\columnwidth]{plots/TTLambdaRawCIWidths90_0_Lambda_SNR}
% \includegraphics[trim={3cm 0 0 0},width=2.\columnwidth]{plots/TTLambdaCIWidths90_0_Lambda_SNR.pdf}
\caption{{\bf Statistical uncertainty in $\lambdans$ measurement:}
Here we show the statistical uncertainty in the measurement of
$\lambdans$, as a percentage of the injected/true value. In each panel, the
same is shown as a function of the BH mass and spin, keeping $\lambdans$ and
injection's SNR $\rho$ fixed (noted in the panel). Rows contain panels
with the same value of $\lambdans$, with $\rho$ increasing from left to right.
Columns contain panels with the same value of $\rho$, with $\lambdans$ 
increasing from top to bottom.
Contours at $(\Delta\lambdans)^{90\%}=\{50\%, 75\%, 100\%, 150\%, 200\%\}\times\lambdans^\mathrm{Injected}$ demarcate regions where we can constrain the
$\lambdans$ parameter well (within a factor of two of the injected value).
We note that, as expected, the measurement accuracy for $\lambdans$ improves
with (i) increasing SNR, (ii) increasing $\lambdans$, (iii) increasing BH spin,
and (iv) decreasing BH mass.
\label{fig:TT_LambdaCIWidths90_0_Lambda_SNR}
% \includegraphics[trim={0 0 0 2cm}, width=0.95\columnwidth]{plots/TTSNRThresholdFor200LambdaMeasurement_BHspin_BHmass_Lambda1500_0_CI90_0}
\includegraphics[width=\columnwidth]{plots/TTSNRThresholdFor100LambdaMeasurement_BHspin_BHmass_Lambda2000_0_CI90_0}
% \includegraphics[width=1.025\columnwidth]{plots/TTSNRThresholdFor100LambdaMeasurement_BHspin_BHmass_Lambda2000_0_CI90_0}\\
We show here, as a function of BH mass and spin, the {\it minimum} signal
strength (SNR) required to constrain $\lambdans$ within an interval of width
equal to $100\%$ of its true value, i.e. with $\pm 50\%$ error-bars. The NS mass
is fixed at $1.35M_\odot$, spin at zero, and $\lambdans=2000$.
We can see that, even in the most conducive circumstances with large aligned 
$\chibh$ and a comparable mass BH, we can only constrain $\lambdans$ to better
than $\pm 50\%$ {\it if} the SNR is $\gtrsim 29$. In the era of design
sensitivity LIGO instruments, we expect this to happen approximately once in a
year of observation~\cite{Abadie:2010cfa}.
% We note that this information is, in principle, contained in 
% Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR}, which we gather here to better understand
% the effect of NS's deformability itself on its measurement. 
% As hinted at in Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR}, we note that the measurability
% of NS matter effects improves with all factors that enhances the signature of the NS's disruption on
% the \textit{detectable} portion of the emitted GW signal (implying, within a frequency band
% set by the detectorTT_SNRThresholds_BHspin_BHmass_CI90_0s).
\label{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0}

@@

% \textbf{Improvement in $\lambdans$ measurement accuracy\\ with multiple observations}\par\medskip
\includegraphics[width=0.8\columnwidth]{plots/pdfLambda_vs_N_L800.pdf}\\
\includegraphics[width=0.8\columnwidth]{plots/FillBetweenErrorBarsLambda_vs_N_L800.pdf}
\caption{{\bf Improvement in $\lambdans$ measurement accuracy with multiple observations:}
These panels show the accumulation of information from a population whose
$\lambdans=800$.
{\it Left:} Posterior probability distributions for $\lambdans$ (colored 
curves), and associated $90\%$ credible intervals (grey vertical lines),
shown for different number of accumulated observations $N$. Distributions are
normalized to unit area.
{\it Right:} Measured median value of $\lambdans$ (as solid circles) and the 
associated $90\%$ credible intervals (as the vertical extent of filled
region), shown as a function of number of observations $N$. Solid horizontal
line indicates the true value of $\lambdans$. Dashed and dotted horizontal
lines (a pair for each line-style) demarcate $\pm 25\%$ and $\pm 50\%$ error
bounds.

@@

% \begin{figure*}
% \centering    
% \includegraphics[width=.9\columnwidth]{plots/FillBetweenErrorBarsLambda_vs_N_L2000.pdf}
% \includegraphics[width=.9\columnwidth]{plots/FillBetweenErrorBarsLambda_vs_N_L1500.pdf}\\
% \includegraphics[width=.67\columnwidth]{plots/FillBetweenErrorBarsLambda_vs_N_L1000.pdf}
% \includegraphics[width=.67\columnwidth]{plots/FillBetweenErrorBarsLambda_vs_N_L800.pdf}
% \includegraphics[width=.67\columnwidth]{plots/FillBetweenErrorBarsLambda_vs_N_L500.pdf}
% \caption{Filled-region plots showing the median and $90\%$ credible intervals
% for $\lambdans$ measurement, as a function of the number of observed events $N$.
% Each panel corresponds a unique population with injected $\lambdans$ fixed to
% the value shown in its title. In each panel, the recovered median for $\lambdans$
% is shown by filled circles, and the $90\%$ credible interval is shown by the 
% height of the filled region. In addition, the dashed and 
% dotted lines (2 of each line-style) show the $\pm 25\%$ and $\pm 50\%$ error
% bounds, with respect to the true value.
% \label{fig:TT_Lambda_vs_N_L500_2000_CI90_0}
% \end{figure*}

@@

% \begin{figure*}
% \centering
% \includegraphics[trim=20 0 0 0, width=1.02\columnwidth]{plots/LambdaMedian_vs_N_AllPopulation}
% \includegraphics[trim=20 0 0 0, width=1.02\columnwidth]{plots/LambdaMedian_vs_N_AstroPopulation}
% \caption{%
% {\it Left:} This figure shows the median value of the recovered
% probability distribution for $\lambdans$, as a function of the number of events
% in the population $N$. There are four family of curves, one corresponding each
% to $\lambdans=\{500,1000,1500,2000\}$, with $100$ independent populations
% within each family. One curve in each family is highlighted in color, 
% representing that it is the same population as was illustrated in
% Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}-\ref{fig:TT_Lambda_vs_N_CI90_0}.
% In the same color we show $\pm 10\%$ error-bounds on $\lambdans$ with
% horizontal dash-dotted lines. We observe that within $10-25$ observations, 
% the median of the measured cumulative probability distribution for $\lambdans$
% will lie within $10\%$ of the true value.
% {\it Right:} This figure is identical to the left panel, with the only
% difference that the BH masses in each population are restricted to lie
% {\it outside} the astrophysical mass-gap (i.e. paradigm B). The difference that
% we observe under this paradigm is that we need more ($30+$) events to achieve 
% the same ($10\%$) measurement accuracy for populations with $\lambdans<1000$.
% For more deformable neutron stars, $10-25$ events would suffice.
% \label{fig:TT_LambdaMedian_vs_N_AllInOne}
% \end{figure*}
% \begin{figure*}
% \centering    
% \includegraphics[width=0.8\columnwidth]{plots/LambdaCIWidths_vs_N_AllPopulations_Log_L2000.pdf}
% \includegraphics[width=0.8\columnwidth]{plots/LambdaCIWidths_vs_N_AstroPopulations_Log_L2000.pdf}\\
% \includegraphics[width=0.8\columnwidth]{plots/LambdaCIWidths_vs_N_AllPopulations_Log_L1500.pdf}
% \includegraphics[width=0.8\columnwidth]{plots/LambdaCIWidths_vs_N_AstroPopulations_Log_L1500.pdf}\\
% \includegraphics[width=0.8\columnwidth]{plots/LambdaCIWidths_vs_N_AllPopulations_Log_L1000.pdf}
% \includegraphics[width=0.8\columnwidth]{plots/LambdaCIWidths_vs_N_AstroPopulations_Log_L1000.pdf}\\
% \includegraphics[width=0.8\columnwidth]{plots/LambdaCIWidths_vs_N_AllPopulations_Log_L800.pdf}
% \includegraphics[width=0.8\columnwidth]{plots/LambdaCIWidths_vs_N_AstroPopulations_Log_L800.pdf}\\
% \includegraphics[width=0.8\columnwidth]{plots/LambdaCIWidths_vs_N_AllPopulations_Log_L500.pdf}
% \includegraphics[width=0.8\columnwidth]{plots/LambdaCIWidths_vs_N_AstroPopulations_Log_L500.pdf}
% \caption{%
% These figures show the width of the $90\%$ credible interval for $\lambdans$
% (normalized by its true value), as a function of the number of observed events
% $N$. The left column panels show populations sampled under paradigm A, which
% allows BH masses to fall within the astrophysical mass-gap, while those on the
% right show populations drawn under paradigm B which respects the mass-gap.
% Each panel corresponds to a unique value of populations' $\lambdans$,
% decreasing from $2000\rightarrow 500$ as we go from top to bottom. In each panel,
% $100$ curves are shown, each corresponding to an independent population draw. The
% one colored curve in each panel highlights the population used in
% Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}-\ref{fig:TT_Lambda_vs_N_CI90_0}.
% The insets in each panel show the same information, with the ordinate {\it not}
% normalized by the true value of $\lambdans$.
% We find that with approximately $25$ or so events, we begin to put
% statistically meaningful constraints on $\lambdans$, restricting it to within
% $\pm 50\%$ of the true value. We can expect to achieve this with a few years
% of design aLIGO operation~\cite{Abadie:2010cfa}. Further tightening of the
% credible intervals will require $40+$ events.
% \label{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne}
% \end{figure*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Acknowledgments
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgments}
  We thank Benjamin Lackey, Francesco Pannarale, Francois Foucart, and Duncan Brown
  for helpful discussions. We gratefully acknowledge support
  for this research at CITA from NSERC of Canada, the Ontario Early 
  Researcher Awards Program, the Canada Research
  Chairs Program, and the Canadian Institute for Advanced Research; at
  Caltech from the Sherman Fairchild Foundation and NSF grants
  PHY-1404569 and AST-1333520; at Cornell from the
  Sherman Fairchild Foundation and NSF grants PHY-1306125 and
  AST-1333129; and at Princeton from NSF grant PHY-1305682 and the
  Simons Foundation.  Calculations were performed at the Vulcan
  supercomputer at the Albert Einstein Institute;
  H.P. and P.K. thank the Albert-Einstein Institute,
  Potsdam, for hospitality during part of the time where this research
  was completed. M.P. thanks CITA for hospitality where part of the work
  was carried out.
\end{acknowledgments}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Techniques}\label{s1:techniques}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% #################
% THESE FIGURES ARE FOR THE NEXT SECTION
\begin{figure*}
\centering 
% \fbox{
% \textbf{Illustrative posterior probability distributions for mass-ratio $\eta$,}\par
% \textbf{at different SNR values (shown in legend).}\par\medskip
\includegraphics[trim=20 18 18 18 0,clip=true,width=1.8\columnwidth]{plots/SingleSystemEta_q4_0_mc2_25_chi0_50}\\
\caption{{\bf Illustrative posterior probability distributions for mass-ratio $\eta$ at different SNR values:}
We show here probability distributions for mass ratio $\eta$ as measured
for the same signal at different SNRs. The intrinsic parameters of the source
are: $q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$, and $\lambdans=2000$;
and the signal is injected at SNRs $\rho=\{20,30,50\}$ (left to right). The templates
{\it ignore} tidal effects.
% 
In each panel: the dashed red line marks the median value
$\eta^\mathrm{Median}$, while the dashed green line show the true value
$\eta^\mathrm{Injected}$. The darker shading shows
the recovered $90\%$ credible interval for $\eta$, $(\Delta\eta)^{90\%}$.
% 
Comparing systematic and statistical errors, we find that:
at $\rho=20$, $\eta$ measurement is dominated by statistical
errors; at $\rho=30$, the two become comparable; and 
for louder signals ($\rho\simeq50$), the systematic errors dominate.
}
% }
\label{fig:SingleSystemEtaPDFvsSNR}
\end{figure*}
% 
% ###############################
% In this section we summarize various technical aspects of this paper. The model
% used to obtain tidal waveforms is described in Sec.~\ref{s2:waveforms}. The 
% process of inferring source parameters from a GW signal is summarized in 
% Sec.~\ref{s2:bayesian}. The process of generating and combining multiple events
% is instead detailed in the corresponding results section,
% Sec.~\ref{s2:astro_multiple}.

@@

@@

\includegraphics[trim=20 20 18 22 0,clip=true,width=1.95\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear}

@@

\includegraphics[trim=20 20 18 21 0,clip=true,width=1.95\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR20_50_linear}

@@

\includegraphics[trim=20 20 18 20 0,clip=true,width=1.95\columnwidth]{plots/TNChiBHBiasesOverCIWidths_CI90_0_Lambda_SNR_linear}

@@

\includegraphics[trim=10 20 8 18 0,clip=true,width=1.9\columnwidth]{plots/SingleSystemLambdaVary_q4_0_mc2_25_chi0_50_snr50}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{What do we gain by using templates that include NS matter effects?}\label{s1:PEwithNS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % 
% #################
\begin{figure*}
\centering    
\includegraphics[trim=35 21 15 21 0,clip=true,width=2.2\columnwidth]{plots/TTLambdaRawCIWidths90_0_Lambda_SNR}
% \includegraphics[trim={3cm 0 0 0},width=2.\columnwidth]{plots/TTLambdaCIWidths90_0_Lambda_SNR.pdf}
\caption{{\bf Statistical uncertainty in $\lambdans$ measurement:}
Here we show the statistical uncertainty in the measurement of
$\lambdans$. In each panel, the
same is shown as a function of the BH mass and spin, keeping $\lambdans$ and
injection's SNR $\rho$ fixed (noted in the panel). Rows contain panels
with the same value of $\lambdans$, with $\rho$ increasing from left to right.
Columns contain panels with the same value of $\rho$, with $\lambdans$ 
increasing from top to bottom.
% 
Contours at $(\Delta\lambdans)^{90\%}=\{50\%, 75\%, 100\%, 150\%, 200\%\}\times\lambdans^\mathrm{Injected}$ demarcate regions where we can constrain the
$\lambdans$ parameter well (within a factor of two of the injected value).
% 
We note that, as expected, the measurement accuracy for $\lambdans$ improves
with (i) increasing SNR, (ii) increasing $\lambdans$, (iii) increasing BH spin,
and (iv) decreasing BH mass.
}
\label{fig:TT_LambdaCIWidths90_0_Lambda_SNR}
\end{figure*}
% % 
\begin{figure}
\centering    
\includegraphics[trim=10 10 0 10 0,clip=true,width=1.05\columnwidth]{plots/TTSNRThresholdFor100LambdaMeasurement_BHspin_BHmass_Lambda2000_0_CI90_0}
% \includegraphics[width=1.025\columnwidth]{plots/TTSNRThresholdFor100LambdaMeasurement_BHspin_BHmass_Lambda2000_0_CI90_0}\\
\caption{
We show here, as a function of BH mass and spin, the {\it minimum} signal
strength (SNR) required to constrain $\lambdans$ within an interval of width
equal to $100\%$ of its true value, i.e. with $\pm 50\%$ error-bars. The NS mass
is fixed at $1.35M_\odot$, spin at zero, and $\lambdans=2000$.
% 
We can see that, even in the most conducive circumstances with large aligned 
$\chibh$ and a comparable mass BH, we can only constrain $\lambdans$ to better
than $\pm 50\%$ {\it if} the SNR is $\gtrsim 29$. In the era of design
sensitivity LIGO instruments, we expect this to happen approximately once in a
year of observation~\cite{Abadie:2010cfa}.
}
\label{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0}
\end{figure}
% %

@@

\includegraphics[trim=18 18 18 10 0,clip=true,width=\columnwidth]{plots/pdfLambda_vs_N_L800.pdf}\\
\includegraphics[trim=18 18 18 10 0,clip=true,width=\columnwidth]{plots/FillBetweenErrorBarsLambda_vs_N_L800.pdf}
\caption{{\bf Recovery of $\lambdans$ for an increasingly large population of BH-NS signals.}
{\it Top}: Posterior probability distributions for $\lambdans$ (colored curves), and
associated $90\%$ credible intervals (grey vertical lines), shown for different number
of accumulated observations N. Distributions are normalized to unit area.  
{\it Bottom}: Measured median value of $\lambdans$ (as solid circles) and the
associated $90\%$ credible intervals (as the vertical extent of filled region), shown as
a function of number of observations N. Solid horizontal line indicates the true value of
$\lambdans=800$. Dashed and dotted horizontal lines (a pair for each line-style) demarcate
$\pm 25\%$ and $\pm 50\%$ error bounds.

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Acknowledgments
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgments}
  We thank Benjamin Lackey, Francesco Pannarale, Francois Foucart, and Duncan Brown
  for helpful discussions. We gratefully acknowledge support
  for this research at CITA from NSERC of Canada, the Ontario Early 
  Researcher Awards Program, the Canada Research
  Chairs Program, and the Canadian Institute for Advanced Research.
  Calculations were performed at the Vulcan
  supercomputer at the Albert Einstein Institute;
  H.P. and P.K. thank the Albert-Einstein Institute,
  Potsdam, for hospitality during part of the time where this research
  was completed. M.P. thanks CITA for hospitality where part of the work
  was carried out.
\end{acknowledgments}
